---
title: "Regression assumes rather than tests causality"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}

# set default chunk options
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

```{r}

# dependencies
library(tidyr)
library(dplyr)
library(purrr) 
library(lavaan)
library(semPlot)
library(knitr)
library(kableExtra)
library(parameters)

```

# Y-causes-X population analyzed with X-causes-Y model

## Plots

Data generating process vs. model used in analysis

```{r fig.height=6, fig.width=7}

# simple effect
mod_1 <- "X ~ Y"

layout_matrix <- matrix(c(-1,  0,
                           1,  0), 
                        ncol = 2, 
                        byrow = TRUE)

lavaan::simulateData(model = mod_1, sample.nobs = 100) %>%
  sem(model = mod_1, data = .) |>
  semPaths(whatLabels = "diagram", 
           layout = layout_matrix, 
           residuals = FALSE,
           edge.label.cex = 1.2, 
           sizeMan = 10)
title(main = "Data generating process in the population:", adj = 0)


# simple effect - reversed direction of causality
mod_2 <- "Y ~ X"

layout_matrix <- matrix(c( 1,  0,
                          -1,  0), 
                        ncol = 2, 
                        byrow = TRUE)

lavaan::simulateData(model = mod_2, sample.nobs = 100) %>%
  sem(model = mod_2, data = .) |>
  semPaths(whatLabels = "diagram", 
           layout = layout_matrix, 
           residuals = FALSE,
           edge.label.cex = 1.2, 
           sizeMan = 10)
title(main = "Model used in analysis:", adj = 0)

```

## Simulate

```{r}

# define data generating process ----
generate_data <- function(population_model, n) {
  
  data <- lavaan::simulateData(model = population_model, sample.nobs = n) |>
    mutate(id = row_number())
    
  return(data)
}

# define analysis ----
# extract the regression estimate for y ~ x
# nb requires that analysis_model includes a y ~ x effect
analyse_yx <- function(data, analysis_model){

  # fit model
  fit <- sem(model = analysis_model, data = data)
  
  # extract results
  results <- fit |>
    parameters() |>
    as_tibble() |>
    filter(To == "Y" & From == "X") |>
    dplyr::select(estimate_yx = Coefficient, p_yx = p)
  
  return(results)
}

# run simulation ----
simulation_yx <- 
  # define simulation conditions
  tibble(n = 100,
         population_model = "X ~ 0.5*Y",
         population_label = "Y causes X",
         population_yx = 0.5,
         analysis_model = "Y ~ X", 
         analysis_label = "X causes Y") |>
  expand_grid(iteration = 1:1000) |>
  # for each one, generate data
  mutate(data = pmap(list(population_model, n), generate_data)) |>
  # for each dataset, analyze it
  mutate(results = pmap(list(data, analysis_model), analyse_yx)) 

# summarize across iterations ----
simulation_summary_yx <- simulation_yx |>
  unnest(results) |>
  group_by(population_label, analysis_label, population_yx) |>
  summarize(mean_estimate_yx = mean(estimate_yx), 
            proportion_significant = mean(p_yx < .05),
            .groups = "drop")

# print table of results
simulation_summary_yx |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

# Y-causes-X population analyzed with Y-causes-X model

## Simulate

```{r}

# define data generating process ----
generate_data <- function(population_model, n) {
  
  data <- lavaan::simulateData(model = population_model, sample.nobs = n) |>
    mutate(id = row_number())
    
  return(data)
}

# define analysis ----
# extract the regression estimate for y ~ x
# nb requires that analysis_model includes a y ~ x effect
analyse_xy <- function(data, analysis_model){

  # fit model
  fit <- sem(model = analysis_model, data = data)
  
  # extract results
  results <- fit |>
    parameters() |>
    as_tibble() |>
    filter(To == "X" & From == "Y") |>
    dplyr::select(estimate_yx = Coefficient, p_yx = p)
  
  return(results)
}

# run simulation ----
simulation_xy <- 
  # define simulation conditions
  tibble(n = 100,
         population_model = "X ~ 0.5*Y",
         population_label = "Y causes X",
         population_yx = 0.5,
         analysis_model = "X ~ Y", 
         analysis_label = "Y causes X") |>
  expand_grid(iteration = 1:1000) |>
  # for each one, generate data
  mutate(data = pmap(list(population_model, n), generate_data)) |>
  # for each dataset, analyze it
  mutate(results = pmap(list(data, analysis_model), analyse_xy)) 

# summarize across iterations ----
simulation_summary_xy <- simulation_xy |>
  unnest(results) |>
  group_by(population_label, analysis_label, population_yx) |>
  summarize(mean_estimate_yx = mean(estimate_yx), 
            proportion_significant = mean(p_yx < .05),
            .groups = "drop")

# print table of results
simulation_summary_xy |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

# Interpret the results

What results are similar or different between the two simulations? What problems could this cause in substantive research?

```{r}

bind_rows(simulation_summary_yx,
          simulation_summary_xy) |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

# Extend your knowledge

We have committed a coding sin in the above: we have repeated ourselves a lot. How could we combine the two simulations above into a single simulation?

# Session info

```{r}

sessionInfo()

```


